---
title: "Outlier Detection"
author: "Michael Koohafkan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Outlier Detection}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette describes functions available for outlier detection,
i.e., the detection of data Values outside some specification based
on the statistical or structural distribution of Values.

## General structure and output

All outlier detection functions follow a similar template of inputs
and outputs. All outlier detection functions accept the following 
arguments:

 1. a vector of data Values;
 2. a logical "mask" used to restrict the calculation of certain
   parameters to a subset of the data; and
 3. A specification of thresholds that discriminate between 
    non-outliers, "mild" outliers, and "extreme" outliers.

All functions return an ordered factor tagging each data Value
as a non-outlier (1), a mild outlier (2), or an extreme outlier (3).
Some outlier detection functions can alternatively return the actual
test statistic or score used to classify the data by specifying the
argument `return.score = TRUE`.

## Parametric approaches

`ODWGtools` provides the following functions for parametric
(distribution-based) outlier detection:

 - `outlier_tscore()`: outlier detection based on the one-sample
   t-test. Note that for sample sizes greater than 30, the t-test
   (t-distribution) approach is equivalent to the z-score 
   (normal distribution) approach.
 - `outlier_chisq()`: outlier detection based on the $\chi^2$
   distribution.

## Non-parametric approaches

`ODWGtools` provides the following functions for non-parametric
(quantile-based) outlier detection:
 
 - `outlier_tukey()`: Tukey's test for outliers (interquartile range).
 - `outlier_mad()`: Outlier detection based on median absolute
   deviation (MAD).

## Other approaches

`ODWGtools` provides the following additional functions 
for outlier detection:
 
 - `outlier_lof()`: Local Outlier Factor (LOF) test for outliers.
 - `outlier_iforest()`: outlier detection based on Isolation Forests.
 - `outlier_glosh()`:  Global-Local Outlier Score from Hierarchies (GLOSH) test for outliers.
 - `outlier_hdbscan()`: outlier detection based on Hierarchical DBSCAN.

## Example Application

We use the [cder package](https://cran.r-project.org/package=cder)
to download a sample of Belden's Landing salinity data
to demonstrate the outlier detection functions.

```{r}
library(cder)
bdl = cdec_query("BDL", 100L, "E", "2018-07-01", "2018-08-01")
```

Univariate outlier detection functions are predicated with
`outlier_`.

```{r}
library(ODWGtools)

bdl["tscore"] = outlier_tscore(bdl$Value)
bdl["tukey"] = outlier_tukey(bdl$Value)
bdl["mad"] = outlier_mad(bdl$Value)
```

Multivariate outlier detection methods are predicated with
`moutlier_`.

```{r}
bdl["chisq"] = moutlier_chisq(bdl[c("DateTime", "Value")])

bdl["lof"] = moutlier_lof(bdl[c("DateTime", "Value")])
bdl["iforest"] = moutlier_iforest(bdl[c("DateTime", "Value")])
```


The [`slider`](https://cran.r-project.org/package=slider) package can
be used to apply `ODWGtools` functions to moving windows. 

```{r}
library(slider)

bdl["tscore.window"] = slide_chr(bdl$Value, ~outlier_tscore(.x)[49], 
  .before = 48, .after = 48)
```

`slider` performs row-wise iteration over dataframes, so the syntax for
using sliding windows with the multivariate outlier detection functions
is straightforward provided you pass a dataframe (rather than a list)).

```{r}
bdl["chisq.window"] = slide_chr(bdl[c("DateTime", "Value")],
  ~ moutlier_chisq(.x)[49],
  .before = 48, .after = 48, .complete = TRUE)

# this won't work
# bdl["chisq.window"] = slide_chr(as.list(bdl[c("DateTime", "Value")]),
#   ~ moutlier_chisq(.x)[49], 
#   .before = 48, .after = 48, .complete = TRUE)
```

Finally, a vignette isn't really complete without some plots.

```{r}
library(ggplot2)

ggplot(bdl) + theme_bw() +
  aes(x = DateTime, y = Value, color = tscore) + 
  geom_point()

ggplot(bdl) +
  aes(x = DateTime, y = Value, color = tscore.window) +
  geom_point()

ggplot(bdl) + aes(x = DateTime, y = Value, color = chisq) + geom_point()

ggplot(bdl) + aes(x = DateTime, y = Value, color = chisq.window) + geom_point()
```